Examining a Quarter Century of Publishing Trends in Social Work Research:

### 2015-05-25 manuscript.Rmd
Major writing update. No code work.

---
title: 'Examining a Quarter Century of Publishing Trends in Social Work Research:
  A Data Science Perspective'
author: ''
output: pdf_document
---

@@

@@

### Journal summary

Figure 1 uses a small multiples plot to summarize the number of article records
for each year throughout the study period.  According to Tufte (xxxx), small multiples (also referred to as
trellis and lattice charts) are the best design solution for visually enforcing
comparisons of change.  In the current useage, each journal is summarized with its own line plot using two different lines to represent the number of article records contained in the SWHD (orange line) and Scopus (purple line).  The journals are sorted in descending order based on the overall number of article records in the SWHD.  Journal titles are also abbreviated, and the full spellings are available in Appendix A.  These line plots help show publication patterns for each journal.

The differences in publication counts are explained by a couple different reasons.  Scopus does not index journals prior to 1996, whereas the SWHD goes back to 1989.  Although the timeframe of the SWHD is longer than Scopus, the actual number of article records in the SWHD for many journals is lower than what is available in Scopus.  Scopus included editorials and book reviews in their article count, whereas these publications were specifically excluded from the SWHD.  Because Scopus data were available only in aggregate form, it was not possible to extract these data to make more refined comparisons.  In the construction of the SWHD, we also applied a series of cleaning procedures that further reduced the article count for many journals.  For example, we excluded article records that did not contain author names, an abstract or article title.  Again, because we were working only with aggregate data from Scopus, we were unable to apply the same cleaning procedures to the Scopus data.  And, finally, differences can also be attributed to gaps and errors in indexing in Scopus and all the databases used to construct the SWHD.  This is a known and longstanding problem with some social work journals (see Holden, Holden).

\newpage
```{r fig.height=10, fig.width=7.5}
plotSmallMultiples.1

A bubble plot was created to facilitate interpretation of these data (see Figure 2).  This involved creating a scatter plot, mapping each journal's total number of publications on the X axis and h-Index value on the Y axis.  Each journal was assigned a quartile rating for its h-Index, and those quartile values were mapped to four unique colors.  Finally, the number of years of representation in the study was mapped to the size of the point, or _bubble_.  Journals without an h-Index rating are not represented in this plot (N = XX).  A smoothed loess line and confidence interval were added to show the trend between the h-Index scores and number of articles published.

The loess trend line shows a moderate association between the h-Index and number of articles published.  However, a number of notable deviations are observed.  Eleven journals were in the highest h-Index quartile.  The _British Journal of Social Work_ had the highest h-Index value but ranked third with regard to the overall number of publications. XXXX, XXXX, and XXXX were also in the top quartile but had comparably fewer published articles and existed for less time than the other journals in the same quartile. _Families in Society_ had the most published articles, spanning the full 25-year study period, but was in the second h-Index quartile.

Figure 3 provides a descriptive summary of the number of authors per article over time.  As indicated by the median, it was more common for articles to be sole-authored until 2006, at which time the median fluctuated between one and two authors, and then stabilizing in 2009 at 2 authors.  The steady increase in the standard deviation since 2000 indicates an increase in the number of authors for each article. The length of articles also increased over time, starting at roughly 12.5 pages in 1989 and increasing steadily over time and stabilizing around 15 pages in 2013.  The standard deviation remained relatively stable over the 25 year period, with all values falling between the range of XX and XX.

@@

>>>>

--
title: 'Examining a Quarter Century of Publishing Trends in Social Work Research:
  A Data Science Perspective'
author: ''
output: pdf_document
---

@@

Why is this study important:

1.  Substantive contributions 
Understand publication trends from a macro perspective.  This can:

From an economic standpoint, the field of social work observed growth
throughout the economic crisis and is expected to maintain growth.  The number
of MSW programs also continues to grow.  While growth is observed in a variety
of areas, it is important to consider the extent to which growth has occurred
regarding overall scholarship.

From a methodological standpoint, this study makes a few unique and important
contributions to the social work research.  Foremost, the study was designed
and executed in a manner that meets the full standards of reproducible
research.  The CTSPEDIA, an advocate of fully reproducible research, defines
this type of research as, ``the practice of distributing, along with a research
publication, all data, software source code, and tools required to reproduce
the results discussed in the publication.  As such the RR [reproducible
research] package not only describes the research and its results, but becomes
a complete laboratory in which the research can be reproduced and extended."
(NOTE ABOUT ORGMODE)

Methods

Include section on reproducible research.

Discussion

## Growth of Social Work Research

From the years 1989 to 2013, the field of social work research produced nearly
35,000 journal articles. This doesn't include books or book chapters,
editorials, book reviews, and other types scientific communications.  The
growth increased steadily from XX to XX, and appears to be leveling off.
Through this period of time, we observed XX unique titles appear while a number
of other journals were lost.

The current study helps clarify the extent to which _team science_ is emerging
in social work research.  This is an important issue to consider, as it has
implications for training researchers, reviewing and understnanding tenure and
promotion cases, and research infrastructure to facilitate collaborations
(http://www.ncbi.nlm.nih.gov/pmc/articles/PMC3321548/). According to Thomson
Reuters, the average number of authors on papers included in the Science
Citation Index increased by about 50% between 1990 and 2010
(http://www.hopkinsmedicine.org/institute_basic_biomedical_sciences/news_events/articles_and_stories/employment/2013_01_middle_authors.html)
In our research, a t least one-half of all published studies were authored by a
single individual until 2009.  The standard deviation continues to increase,
providing further evidence of continued team science growth in social work.
How authorship is assessed is a very important consideration for the field,
particularly with respect to tenure reviews.  It is not clear how reviewers of
tenure cases are viewing team science contributions and _middle authors_.
However, this is something that will have to be considered for the field. In
the field of physics, the weighted average seniourity and average seniority
measures both had a negative relationship with the impact, based on the number
of citations the publication received, as well as a negative relationship
between the number of authors and the number of citations.  http://myweb.fsu.edu/bstvilia/papers/AuthorTeamDiversityImpact.pdf
(***NOTE TO BRYAN -- AS A QUICK STUDY, YOU AND I SHOULD BUST OUT A STUDY
INVOLVING THIS.  WE CAN DO IT FOR JUST THE TOP JOURNALS IN SOCIAL WORK THAT ARE
IN WOS.)

http://dmm.biologists.org/content/7/3/311.full

https://ccrod.cancer.gov/confluence/download/attachments/47284665/teamscience_fieldguide.pdf

One thing to consider is the extent to which social work research is being
served by other journal outlets.  For example, social work researchers with
primary research interests in areas of health, mental health, and substance use
may publish in speciality journals outside of social work for a variety of
reasons.  Thus, the amount that social work has grown should be cautious,
keeping in mind this particular metric refers only to social work journals.
The field may have grown more rapidly in other areas that are not reflected in
the current research.  The functions that were constructed to _wrangle_ the
abstracts are functional for records extracted from PubMed.  Since this data
source contains author affiliations, it would be feasible to extract this
information using a set of regular expressions.

According to the journal Nature, from the lat 1600s until about 1920, the rule
was one author per paper.  The rules was ``breached in the 1920s, diminished in
the 1950s, and largely abandoned by the 1980s'' (p.
http://www.nature.com/nature/history/full/nature06243.html).  Social work
research had a much different trajectory.  In social work research, the rule
wasn't breached until 2009.

This has implications for tenure, and it is unclear how _team science_ is
actually being considered in context of these reviews.  The field is arguably
in a time of transition, form si

Defining a _disciplinary_ social work journal was a complicated task.  Social
work is inherently inter-disciplinary.  We based our definition of a social
work journal using the list of Hodge and Lacasse (2011), who created a
comprehensive list of journal titles.  This list was based on earlier attempts
to identify all the disciplinary social work journals.  We supplemented our
list by doing supplemental searches in all data sources for journals that used
any of the following terms.

The length of the articles was examined, as it has implications for device
formatting.  Clearly, the field of social work has a current average of 15
typeset pages.  This consistency allows for targeted customization for mobile
devices.  An article can be optimized and fairly standardized, which has
important benefits for information retrieval and comprehension.

This article raised many opportunities that became obvious in this study.  In
the current study, we worked in earnest to present as much of the analyses
visually.  At the time of writing this article, the majority of social work
journals do not have full color articles.  Indeed, this is a residual policy
from when journals were exclusively in black-and-white and before electronic
archiving.  Now, the majority of research is disseminated electronically, and
even the most primitive computers connected to the Internet have color
monitors.

Data quality -- is there anything to talk about on this topic???

Team science -- is it good for social work research?

## Reproducible research

Reproducible research is essential to ensuring the integrity of the scientific
record.  Making research reproducible is a complicated process that does carry
risks.  More specifically, it does involve taking extra time to prepare
documentation, organize code, and prepare files to be made publicly available.
This problem has already been given careful attention in other fields, which
can be considered reasonable solutions for solution work research.  For
example, this study utilized Rmarkdown as the tool for producing reproducible
research.  Rmarkdown is ... [provide *short* description here -- tell reader
minimum she/he needs to know].

Another practical consideration of reproducible research is curating the data,
and making it available.  The availability can be challenging.  For example, 
many social work studies involve sensitive data that ultimately requires very
strict limitations on how and where the data are stored, along with
restrictions on who can access the data.  [Describe the PS journal that helps
address this issue ...]  This study involves proprietary data from EbscoHost
and ProQuest, which prevents us from making all the data available.  However,
we provide a minimum random extract of the data to ensure verification of our
computer code.  Additionally, we have made all aspects of our computer code
available that will allow researchers with access to EbscoHost and ProQuest to
replicate our findings, as well as use the code to build upon this research.

We believe that social work journals should consider implementing policies that
encourage reproducible research.  Making this best practice in publishing
represents a straightforward way to improve the quality of social work
research.  Along with a policy of reproducible research, the issue of how the
journal responds to errors will also be necessary.  We certainly do not
advocate that authors are expected to retract articles or face public shame in
the context of random human error.  Some studies may involve generating
thousands of lines of computer code -- and, a simple bug in the code can
produce erroneous results.  Researchers who adhere to the principles of
reproducible research are ultimately working in earnest to ensure they are
producing the highest quality research possible.  Errors that are identified in
the process represent an opportunity to further advance knowledge, which
sometimes means correcting knowledge.  These situations simply reveal that the
scientific method is actually working the way it is supposed to.  At present,
we have to assume that all the social work research published to date is
without error.  In absence of reproducible research, we can not fully assess
the quality of research or replicate a study.

Discussion point:

Another complication is the possible consequences that a researcher might experience
when an error is discovered.  This is an important issue that the field needs
to address.

very research report is subject to some form of human
error.  It is important to consider some of the different types of error and
how the field should respond.

(Conduct and analysis on reproducible research" and "replication" or
"replicate")

(original citation from orgmode.org:
http://orgmode.org/worg/org-contrib/babel/intro.html)

 conforms to all
aspects of what is regarded as reproducible research by making available all
the code produced in order to conduct the study.

.  Methodological contributions

@@

### Aggregated summary (Figure 1a. 1b. -- line charts)

Figures 1a and 1b are line charts showing the number of unique journal
titles and journal articles published each year during the study time frame.
In 1989, XX unique journals published a XX articles.  The number of journal
titles showed some variability from 1989 to 1996, followed by nine years of
steady growth.  By 2005, the number of journal titles and journal articles
being published was more than double than what was observed in 1989. The years
2005 to 2013 exhibited some variability in the number of titles and articles
published.  The maximum number of journal titles was observed in 2011 (n = XX)
and dropped to XX by 2013.  The maximum number of journal articles published
was in 2012, and this dropped to XX in 2013.  The cumulative number of journal
articles published over the study time frame was XX,XXX.

### Dis-aggregated summary (Figure 2. Small multiples)

Figure 2 is a small multiples plot, also referred to as a trellis or lattice
chart, that dis-aggregates Figures 1a and 1b.  More specifically, the small
multiples plot shows the number of articles published each year for each 
journal throughout the study time frame (orange line).  We also included data from Scopus for
comparative purposes (purple line).  The small multiples are sorted in descending
order based on the total number of article records in the HD.  The presentation
of the sort is by row, so the overall plot is read from left to right.
Abbreviations are used in Figure 2, and the full journal name is presented in
Appendix A.

The small multiples plot is particularly useful to show the variability in
publishing for each journal.  In fact, according to Tufte (xxxx), small
multiples are the best design solution for visually enforcing comparisons of
change.  The plot shows that _Families in Society_, _Social Work_, and _British
Journal of Social Work_ published the greatest number of articles throughout
the study time frame.  However, the number of articles published by _Families
in Society_ and _Social Work_ appear to be decreasing, while other journals
show an increase -- e.g., _Research on Social Work Practice_, _International
Social Work_, and _Journal of Human Behavior and the Social Environment_.

Data from Scopus was used to help show the overall coverage of the HD, in
addition to revealing possible gaps in indexing.  The HD and Scopus plotted
values do not show exact alignment for a couple reasons.  More specifically,
the HD contains article records going back to 1989, whereas Scopus provides
data starting from 1996.  Additionally, the overall journal coverage of HD is
much greater than Scopus -- that is, Scopus provides data for only xx% of the
journal titles contained in the HD.

The Scopus data that are available show a high degree of correspondence with
the HD.  Small discrepancies between the Scopus and HD can be attributed to
differences in indexing.  That is, the article record count of Scopus includes
editorials and book reviews, whereas these were excluded from the HD.  In the
construction of the HD, we applied a series of cleaning procedures (refer to
Methods section) that further reduced the article count.  Because Scopus data
were available only in aggregate form, it was not possible to extract these
data to make more refined comparisons.

And, finally, the small multiples plot is especially useful for identifying
potential gaps in indexing, which is a problem that has been previously
identified with respect to social work journals (see Holden; Holden). For
example, Scopus appears to be missing article records for the _Journal of
Gerontological Social Work_ from XX to XX.  And, the HD is missing records for
various journals including (but not limited to) the _Journal of Ethnic and
Cultural Diversity in Social Work_, _Journal of Social Work Disability &
Rehabilitation_.

### Article attributes

Figures 3a and 3b provide summaries of two important article attributes -- that
is, the number of authors per article and the length of each article.
Throughout the history of social work research, at least half of all social
work articles were sole authored until 2008, as indicated by the median value
in Figure 3a.  Since 2008, at least half of all articles had two authors.
Throughout the study time frame, the number of authors has increased steadily,
as shown by the increasing mean value and its standard deviation.  In 2013, the
mean number of authors was approximately 2.25, with a standard deviation of
1.4.  The length of the articles increased steadily from approximately 12 pages
in 1989 to 15 pages in 2004.  The last three years of the study time frame
shows a fairly stable publishing trend, with half of all articles being 15
pages (Mean = XX, sd = XX).

### Abstract analysis

The final set of analyses involved text analysis of the article abstracts to
determine whether trends in publishing could be readily identified based on the
frequencies of specific words contained in article abstracts.  In this analysis, we selected words that
represented different _levels of evidence_ within the evidence-based practices
framework.  EBP was selected as the focus of analysis, given that the topic
area is a relatively new in social work, compared to other topics like child
welfare and case management.  Moreover, because EBP has received considerable
attention in various areas of social work research, it was expected that
the different _levels of evidence_ would show should both absolute and
relative changes in their occurrences.

To identify the terms

it is relatively new and has
become part of social work terminology,

 that are common to the framework of
evidence-based practices (EBP).  This topic area was selected because it is a
more recent area of research relative to other social work topics (e.g., child
welfare and case management).

The final part of the study involved a topical analysis by using regular
expressions to identify articles using keywords based on different levels of
evidence.  We specified four sets of keywords relating to meta-analysis,
systematic reviews, randomized controlled trials, and quasi-experimental
studies.  We refer to these as _sets_ of keywords because we used a number of
different variations on the spelling and related terminology.  For example, RCT
involved a search specification that such as "randomized control trial,"
"randomized controlled trial," and "RCT".  The complete definitions of these
are provided in the index.

Figure 4 is a stacked line chart showing the number of each type of article
throughout the study timeframe.  This figure shows increased numbers of all
study types.  It is important to consider this growth relative to the overall
growth of the field of social work.  Thus, in Figure 5, we plotted the
proportion of each keyword class relative to the overall number of articles for
that given year.  This shows an increase -- albeit a small increase -- in the
proportion of journals, providing evidence of growth relative to the overall
field.  Slightly more than 1% of all articles published contained a phrase
equivalent to randomized control trial.

@@

```
\newpage
```{r fig.height=10, fig.width=7.5}
plotSmallMultiples.1
